Accelerating Neural Network Training using Arbitrary Precision Approximating Matrix Multiplication Algorithms

### 2021-04-24 latex/ICPP21/paper.tex
update figures in paper
submitted version to ICPP with VGG 442 data

@@

\begin{abstract}
Matrix multiplication is one of the bottleneck computations for training the weights within deep neural networks.
To speed up the training phase, we propose to use faster algorithms for matrix multiplication known as Arbitrary Precision Approximating (APA) algorithms.
APA algorithms perform asymptotically fewer arithmetic operations than the classical algorithm, but they compute an approximate result with an error that can be made arbitrarily small in exact arithmetic.
Practical APA algorithms provide significant reduction in computation time and still provide enough accuracy for many applications like neural network training.
We demonstrate that APA algorithms can be efficiently implemented and parallelized for multicore CPUs to obtain up to 28\% and 21\% speedups over the fastest implementation of the classical algorithm using one core and 12 cores, respectively.
Furthermore, using these algorithms to train a Multi-Layer Perceptron (MLP) network yields no significant reduction in the training or testing error.
Our performance results on a large MLP network show overall sequential and multithreaded performance improvements of up to 25\% and 13\%, respectively.
\RZ{We also demonstrate that APA algorithms are efficiently implemented on a more complicated network VGG-11?}
\end{abstract}

@@

Our contribution is the use of APA matrix multiplication algorithms to address this problem.
In particular we
\begin{enumerate}
	\item curate a collection of both well-established and recently discovered practical APA algorithms;
	\item extend the framework of \cite{BB15} to generate efficient multithreaded code for all of them, achieving up to 21\% performance improvement over the best parallel classical implementation;
	\item demonstrate the robustness of learning accuracy to approximate matrix multiplications;
	\item present multithreaded performance improvements of a synthetic MLP of up to 13\% over the use of classical matrix multiplication; and
	\item \RZ{present performance improvements of VGG-19 network of 15\% over the classical matrix multiplication.}
\end{enumerate}

@@

\subsection{Skewed Matrix Multiplication}
When computing square matrix multiplications, we have demonstrated that $\dims{4,4,2}$ and $\dims{4,4,4}$ algorithms have the best performance.
However, algorithms with more skewed aspect ratios will likely perform better for problems with matching skewed matrix multiplication dimensions.
We note that an algorithm for dimensions $\dims{m,n,k}$ can be translated into an algorithm for $\dims{n,m,k}$ and any other reordering of the dimensions \cite{BB15}.
\RZ{transposing row major matrix problem?}
\RZ{Best algorithm for each shape of matrices and its figure}
In order to achieve the best performance on a neural network model VGG-11, which we will apply these APA algorithms on, we decided to measure the performance all 28 algorithms on the shapes of multiplied matrices involved in the fully connected layers of VGG-11.
Each fully connected layers of VGG-11 contains the number of weights of 4096, 4096, and 1000, and the input dimension of the first layer is 25088.
Therefore, there are 9 in total multiplications involved: $\dims{bs, 25088, 4096}$, $\dims{bs, 4096, 25088}$, $\dims{25088, bs, 4096}$, $\dims{bs, 4096, 4096}$, $\dims{bs, 4096, 4096}$, $\dims{4096, dims, 4096}$, $\dims{bs, 4096, 1000}$, $\dims{bs, 1000, 4096}$, and $\dims{4096, bs, 1000}$.
($bs$ stands for the batch size of the input.)

The performance results for 6 threads shows that algorithm $\dims{4,4,2}$ and its variances are still the dominant one among all 12 APA algorithms and their variances.

@@

\begin{figure*}
%
\subfloat[One thread]{
\centering
\input{plots/VGG_time_seq}
\label{fig:vgg_seq}}
	%
\hfill	
\subfloat[Six threads]{
\centering
\input{plots/VGG_time}
\label{fig:vgg_par6}}
%
\caption{Total time test on the fully connected layes on VGG-19 using $\dims{4,4,2}$ and dgemm}
\label{fig:vgg}
\end{figure*}

\subsection{VGG-19 Network}
VGG19 is a convolutional neural network typically trained on the imagenet dataset as a classification task. The model is one of the highest performing algorithms for this task. 
The network is composed of 19 layers, 16 convolutional and 3 fully connected. 
The 3 fully connected layers consist of 25,088, 4096 and 1000 nodes and are placed at the end of the network. 
These create very large multiplications in forwards and backwards propagation. 
Using faster APA algorithms for the multiplication we are able to speed up the training of the network.

% \subsection{Accuracy}

\subsection{Performance}

Replacing the classic matrix multiplication algorithm in these layers with the Smirnov 442 algorithm provides up to a 10\% speed up of the fully connected layers during training with 6 threads and a 15\% speed up using a single thread. 
\cref{fig:vgg} plot the training time of each network with increasing batch sizes

@@

As illustrated by Benson and Ballard \cite{BB15}, the highest performing fast algorithms for rectangular matrix multiplications often have dimensions that match the aspect ratio of the problem.
We consider in this paper only square matrix multiplications, and we see that square (or nearly square) dimensions of $\dims{4,4,4}$ and $\dims{4,4,2}$ are fastest.
Algorithms with more skewed aspect ratios will likely perform better for problems with matching skewed matrix multiplication dimensions.
We note that an algorithm for dimensions $\dims{m,n,k}$ can be translated into an algorithm for $\dims{n,m,k}$ and any other reordering of the dimensions \cite{BB15}.
For very large matrices, it may also be beneficial to consider more than one recursive step of the same algorithm or a combination of two or three different algorithms across recursive steps (uniform, non-stationary algorithms \cite{BBDLS16}).

>>>>

\setcopyright{rightsretained}

@@

\begin{abstract}
Matrix multiplication is one of the bottleneck computations for training the weights within deep neural networks.
To speed up the training phase, we propose to use faster algorithms for matrix multiplication known as Arbitrary Precision Approximating (APA) algorithms.
APA algorithms perform asymptotically fewer arithmetic operations than the classical algorithm, but they compute an approximate result with an error that can be made arbitrarily small in exact arithmetic.
Practical APA algorithms provide significant reduction in computation time and still provide enough accuracy for many applications like neural network training.
We demonstrate that APA algorithms can be efficiently implemented and parallelized for multicore CPUs to obtain up to 28\% and 21\% speedups over the fastest implementation of the classical algorithm using one core and 12 cores, respectively.
Furthermore, using these algorithms to train a Multi-Layer Perceptron (MLP) network yields no significant reduction in the training or testing error.
Our performance results on a large MLP network show overall sequential and multithreaded performance improvements of up to 25\% and 13\%, respectively.
We also demonstrate up to 15\% improvement when training the fully connected layers of the VGG-19 image classification network.
\end{abstract}

@@

Our contribution is the use of APA matrix multiplication algorithms to address this problem.
In particular we
\begin{enumerate}
	\item curate a collection of both well-established and recently discovered practical APA algorithms;
	\item extend the framework of \cite{BB15} to generate efficient multithreaded code for all of them, achieving up to 21\% performance improvement over the best parallel classical implementation;
	\item demonstrate the robustness of learning accuracy to approximate matrix multiplications;
	\item present multithreaded performance improvements of a synthetic MLP of up to 13\% over the classical algorithm; and
	\item show performance improvements for training the fully connected layers of the VGG-19 network of up to 15\%.
\end{enumerate}

@@

\begin{comment}
\subsection{Skewed Matrix Multiplication}
When computing square matrix multiplications, we have demonstrated that $\dims{4,4,2}$ and $\dims{4,4,4}$ algorithms have the best performance.
However, algorithms with more skewed aspect ratios will likely perform better for problems with matching skewed matrix multiplication dimensions.
We note that an algorithm for dimensions $\dims{m,n,k}$ can be translated into an algorithm for $\dims{n,m,k}$ and any other reordering of the dimensions \cite{BB15}.
\RZ{transposing row major matrix problem?}
\RZ{Best algorithm for each shape of matrices and its figure}
In order to achieve the best performance on a neural network model VGG-11, which we will apply these APA algorithms on, we decided to measure the performance all 28 algorithms on the shapes of multiplied matrices involved in the fully connected layers of VGG-11.
Each fully connected layers of VGG-11 contains the number of weights of 4096, 4096, and 1000, and the input dimension of the first layer is 25088.
Therefore, there are 9 in total multiplications involved: $\dims{bs, 25088, 4096}$, $\dims{bs, 4096, 25088}$, $\dims{25088, bs, 4096}$, $\dims{bs, 4096, 4096}$, $\dims{bs, 4096, 4096}$, $\dims{4096, dims, 4096}$, $\dims{bs, 4096, 1000}$, $\dims{bs, 1000, 4096}$, and $\dims{4096, bs, 1000}$.
($bs$ stands for the batch size of the input.)

The performance results for 6 threads shows that algorithm $\dims{4,4,2}$ and its variances are still the dominant one among all 12 APA algorithms and their variances.
\end{comment}

@@

\begin{figure}
%
\subfloat[One thread]{
\centering
\input{plots/VGG_time_seq}
\label{fig:vgg_seq}}%
\hfill	
\subfloat[Six threads]{
\centering
\input{plots/VGG_time}
\label{fig:vgg_par6}}
%
\caption{Per-batch training time on the fully connected layers of VGG-19 using $\dims{4,4,2}$ and Classical}
\label{fig:vgg}
\end{figure}

%\subsection{VGG-19 Network}
VGG-19 \cite{SZ15} is a convolutional neural network typically trained on the ImageNet dataset as a classification task. 
The model is one of the highest performing algorithms for this task, and the network is composed of 19 layers, 16 convolutional and 3 fully connected. 
The 3 fully connected layers consist of 25{,}088, 4096, and 1000 nodes, creating very large matrix multiplications in forward and backward propagation. 
By using faster APA algorithms for the multiplication, we are able to speed up the training of the network.

%\subsection{Performance}

Replacing the classic matrix multiplication algorithm in these layers with the $\dims{4,4,2}$ algorithm provides up to a 10\% speed up of the fully connected layers during training with 6 threads and a 15\% speed up using a single thread. 
\Cref{fig:vgg} shows the per-batch training time of the fully connected layers across batch sizes.

@@

As illustrated by Benson and Ballard \cite{BB15}, the highest performing fast algorithms for rectangular matrix multiplications often have dimensions that match the aspect ratio of the problem.
We consider in this paper only square matrix multiplications, and we see that square (or nearly square) dimensions of $\dims{4,4,4}$ and $\dims{4,4,2}$ are fastest.
Algorithms with more skewed aspect ratios will likely perform better for problems with matching skewed matrix multiplication dimensions.
We note that an algorithm for dimensions $\dims{m,n,k}$ can be translated into an algorithm for $\dims{n,m,k}$ and any other reordering of the dimensions \cite{BB15}.
For very large matrices, it may be worth considering more than one recursive step of the same algorithm or a combination of two or three different algorithms across recursive steps (uniform, non-stationary algorithms \cite{BBDLS16}).
